The Impact of Alternation

### 2017-05-12 draft.tex
ok

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig}$ consists of two distinct sorts
$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
\in \ssorts{\sig}$. If $\#(f)=0$ then $f$ is a \emph{constant}. We
consider the constants $\top$ and $\bot$ of sort $\Bool$.

@@

We denote by $\fv{S}{\phi}$ the set of free (i.e.\ not under the scope
of an existential quantifier) variables of sort $S$ in $\phi$ and
write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$.
%
%% A formula is in positive normal form (PNF) if negation occurs only
%% within literals and in disjunctive normal form (DNF) if it is a
%% disjunction of conjunctions of literals $\bigvee_{i=1}^N
%% \bigwedge_{j=1}^{M_i} \ell_{ij}$.
%
The size of a formula $\phi$, denoted as $\len{\phi}$, is the number
of symbols occuring in it.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$.

@@

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta(\phi)$ and, respectively,
$\Delta^k(\phi,a)$, for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$, respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k > 0$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ by induction hypothesis} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ by the definition of $\overline{\phi}$} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, the antichain-based semi-algorithm diverges in a large
number of cases, since infinite antichains exist in many
interpretations, e.g.\ $q \wedge x \teq 0,~ q \wedge x \teq 1, \ldots$
for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (\ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (\ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
\emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

The solution we adopt is abstraction to ensure the termination of
invariant computations. However, it is worth pointing out from the
start that abstraction alone will only allow us to build invariants
that are not necessarily safety invariants. To meet the latter
condition, we resort to counterexample guided abstraction refinement
(CEGAR).

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can now occur negated within the formulae $\phi \in \preds$. We fix
$\preds$ from now on and refer to these formulae as
\emph{predicates}. Given a formula $\phi$, we denote by $\abs{\phi}
\equiv \bigwedge \set{\pi \in \preds \mid \phi \models \pi}$ the
abstraction of $\phi$ w.r.t. the predicates in $\preds$. The abstract
version of the post-image and acceptance condition are defined as
follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

We are left with fulfilling the second point from the definition of a
safety invariant. To this end, suppose that, for a given set $\preds$
of predicates, the invariant $\abs{\Inv}$, defined by the previous
lemma, meets point (\ref{it1:inv}) but not point (\ref{it2:inv}) from
the definition of invariants, where $\Post{\A}$ and $\Accept{\A}$
replace $\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In
other words, there exists a finite sequence $u \in \Sigma^*$ such that
$\nu \in \abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models
\AbsAccept{\A}(u)$, for some boolean $\beta : Q \rightarrow \booli$
and data $\nu : \vec{x} \rightarrow \Data^\I$ valuations. Such
a $u\in\Sigma^*$ is called a \emph{counterexample}.

Once a counterexample $u$ is discovered, there are two
possibilities. Either\begin{inparaenum}[(i)]
\item $\Accept{\A}(u)$ is satisfiable, in which case $L(\A) \neq \emptyset$, or
\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is called
  \emph{spurious}.
\end{inparaenum}
In the first case, our semi-algorithm stops and returns a witness for
non-emptiness, obtained from the satisfying valuation of
$\Accept{\A}(u)$ and in the second case, we must strenghten the
invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
that $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
is carried out by adding to $\preds$ several predicates that are
sufficient to exclude the spurious counterexample.

>>>>

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
$\top$ and $\bot$ of sort $\Bool$.

@@

We denote by $\fv{\sigma}{\phi}$ the set of free (not under the scope
of an existential quantifier) variables of sort $\sigma$ in $\phi$ and
write $\fv{}{\phi}$ for $\bigcup_{\sigma \in \ssorts{\sig}}
\fv{\sigma}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$
such that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i\in[1,n]$. The size $\len{\phi}$ is the number of symbols occuring
in $\phi$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$. A decision
procedure for $\theory(\sig,\I)$ is an algorithm that takes a formula
$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
\theory(\sig,\I)$.

@@

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta(\phi,a)$ (respectively,
$\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$ (respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k\in[1,n]$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words from
$\Sigma[\vec{x}]^*$ accepted by $\A$.
%
In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
  \setminus L(\A_1)$. \qed}

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
for $u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, even with this optimization, the enumeration of
sequences from $\Sigma^*$ diverges in many real cases, because
infinite antichains exist in many interpretations, e.g.\ $q \wedge x
\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
\emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

The solution we adopt in the sequel is abstraction to ensure the
termination of invariant computations. However, it is worth pointing
out from the start that abstraction alone will only allow us to build
invariants that are not necessarily safety invariants. To meet the
latter condition, we resort to counterexample guided abstraction
refinement (CEGAR).

Formally, we fix a set of formulae $\preds \subseteq
\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
formulae as \emph{predicates}. Given a formula $\phi$, we denote by
$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
$\preds$. The abstract versions of the post-image and acceptance
condition are defined as follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

We are left with fulfilling point (\ref{it2:inv}) from the definition
of a safety invariant. To this end, suppose that, for a given set
$\preds$ of predicates, the invariant $\abs{\Inv}$, defined by the
previous lemma, meets point (\ref{it1:inv}) but not point
(\ref{it2:inv}), where $\Post{\A}$ and $\Accept{\A}$ replace
$\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In other words,
there exists a finite sequence $u \in \Sigma^*$ such that $\nu \in
\abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
for some boolean $\beta : Q \rightarrow \booli$ and data $\nu :
\vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
called a \emph{counterexample}.

Once a counterexample $u$ is discovered, there are two
possibilities. Either\begin{inparaenum}[(i)]
\item $\Accept{\A}(u)$ is satisfiable, in which case $u$ is
  \emph{feasible} and $L(\A) \neq \emptyset$, or
%
\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is
  \emph{spurious}.
\end{inparaenum}
In the first case, our semi-algorithm stops and returns a witness for
non-emptiness, obtained from the satisfying valuation of
$\Accept{\A}(u)$ and in the second case, we must strenghten the
invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
that $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
is carried out by adding to $\preds$ several predicates that are
sufficient to exclude the spurious counterexample.
